---
title: "MEGENA module conservation analysis between cohort 1 and 2"
author: "Thanneer Perumal"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
### Load libraries
```{r libs, echo=FALSE, warning=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(plyr)
library(tidyverse)
library(stringr)

library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

library(foreach)
library(doParallel)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```
### Get synapse and github specific parameters
```{r synapse.parameters, include=FALSE}
# Synapse parameters
PARENT_ID = 'syn11638075'

# Github commit for the source code
thisFileName <- 'moduleConservationAnalysisC1andC2.Rmd'
thisRepo <- getRepo(repository = "th1vairam/CRANIO", ref="branch", refName='moduleAnalysis')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=thisFileName)
```
```{r utility.functions}
modPreservStats <- function(refCoexpMat, testCoexpMat, refModule){
  refModule = intersect(colnames(refCoexpMat), colnames(testCoexpMat)) %>%
    intersect(refModule)
  modLen = length(refModule)
  
  # Extract sub networks from reference and test
  refSubnet = refCoexpMat[unique(refModule), unique(refModule)]
  testSubnet = testCoexpMat[unique(refModule), unique(refModule)]
  
  # Calculate node degree in ref and test networks
  testAllNodeDegree = rowSums(testCoexpMat, na.rm = T)
  testNodeDegree = rowSums(testSubnet, na.rm = T)
  refNodeDegree = rowSums(refSubnet, na.rm = T)
  
  ### Connectivity preservation stats
  # Correlation between adjacency matrices
  cor.adj = cor(as.vector(refSubnet), as.vector(testSubnet)) %>% as.numeric()
  
  # Correlation between node degree
  cor.kIM = cor(refNodeDegree, testNodeDegree[names(refNodeDegree)])

  ### Density preservation stats
  # Mean correlation
  mean.Cor = rowMeans(testSubnet) %>% mean()
  
  # Mean connectivity
  mean.kIM = mean(testNodeDegree)
  
  # Mean connectivity ratio 
  mean.kIM2kALL = mean(testNodeDegree/testAllNodeDegree[names(testNodeDegree)], na.rm = T)
  
  return(data.frame(cor.adj = cor.adj,
                    cor.kIM = cor.kIM,
                    mean.Cor = mean.Cor,
                    mean.kIM = mean.kIM,
                    mean.kIM2kALL = mean.kIM2kALL))
}
```
### Get modules and coexpression matrix from synapse
```{r get.mdls}
mod = c(cohort1 = 'syn11635629', cohort2 = 'syn11635905') %>%
  plyr::llply(.fun =function(foldId){
   synQuery(paste0('select * from folder where parentId =="', foldId, '"')) %>%
      plyr::dlply(.(folder.name), .fun = function(x){
        tmp1 = synQuery(paste0('select * from file where parentId =="', x$folder.id, '"')) %>% 
          dplyr::filter(file.name == 'output.RData')
        load(synGet(as.character(tmp1$file.id))@filePath)
        
        # Remove small or big modules
        mod = output$modules
        sz = sapply(mod, length)
        mod = mod[sz >= 30 & sz <= 2000]
        return(mod)
      })
  })

total.mod = plyr::ldply(mod, function(x){
  plyr::ldply(x, function(y){
    data.frame(nmod = length(y),
               median.sz = median(sapply(y, length)))
  })
}) %>%
  dplyr::rename(cohort = .id, subtype = folder.name) %>%
  tidyr::gather(feature, value, -cohort, -subtype) %>%
  tidyr::unite(cohort.feature, cohort, feature, sep = '.') %>%
  tidyr::spread(cohort.feature, value)
kable(total.mod)
  
cor.mat.ids = c(cohort1 = 'syn11635629', cohort2 = 'syn11635905') %>%
  plyr::llply(.fun =function(foldId){
   synQuery(paste0('select * from folder where parentId =="', foldId, '"')) %>%
      plyr::dlply(.(folder.name), .fun = function(x){
        tmp1 = synQuery(paste0('select * from file where parentId =="', x$folder.id, '"')) %>% 
          dplyr::filter(file.name == 'Data_Correlation.txt') %>%
          dplyr::select(file.id)%>%
          unlist() %>%
          as.character()
       
      })
  })
```
### Get module overlap/enrichment statistics
```{r ovlp.megena}
bck.genes = unique(c(do.call(c, mod$cohort1$control), 
                     do.call(c, mod$cohort2$control)))
ovlp.stats = purrr::map2(mod$cohort1, mod$cohort2, function(c1, c2, bckGenes){
  plyr::ldply(c1, .fun = function(x, c2, bckGenes){
    plyr::ldply(c2, .fun = function(y, x, bckGenes){
      CovariateAnalysis::fisherEnrichment(x,y,bckGenes)
    }, x, bckGenes, .parallel = T, .id = 'ModuleName_c2')
  }, c2, bckGenes, .id = 'ModuleName_c1') %>%
    dplyr::mutate(FDR = p.adjust(pval, method = 'fdr')) %>%
    dplyr::filter(FDR <= 0.05, Odds.Ratio >= 1)
}, bck.genes)

ovlp.mod = ovlp.stats %>%
  purrr::map(.f = function(x){
    x = x %>% 
      dplyr::filter(FDR <= 0.05, noverlap >= 15) %>%
      dplyr::select(ModuleName_c1, ModuleName_c2, Odds.Ratio) %>%
      dplyr::mutate(ModuleName_c1 = paste0('cohort1.', ModuleName_c1),
                    ModuleName_c2 = paste0('cohort2.', ModuleName_c2)) %>%
      igraph::graph_from_data_frame(directed = F) %>%
      igraph::as_adjacency_matrix(attr = 'Odds.Ratio') %>%
      as.matrix()
    return(list(cohort1 = grep('cohort1', rownames(x), value = T) %>% gsub('cohort1.','',.),
                cohort2 = grep('cohort2', rownames(x), value = T) %>% gsub('cohort2.','',.)))
  })
```
### Get module conservation statistics across cohort 1 and 2 in each sub-type
```{r mdl.cnsrv}
conservation.stats.cohort = purrr::transpose(mod) %>%
  purrr::imap_dfr(.f = function(innerMod, subType, corMatIds){
  coexpMat = c(cohort1 = corMatIds$cohort1[[subType]], cohort2 = corMatIds$cohort2[[subType]]) %>%
    plyr::llply(.fun = function(id){
        synGet(as.character(id))@filePath %>%
          fread(data.table = FALSE, header = T) %>%
          igraph::graph_from_data_frame(directed = FALSE) %>%
          igraph::as_adjacency_matrix(attr = 'rho', type = 'both') %>%
          as.matrix() %>%
          return()
      })
  
  consvStat = purrr::imap_dfr(innerMod, .f = function(modList, cohortName, coexpMat){
    
    refCoexpMat = purrr::keep(coexpMat, names(coexpMat) == cohortName)[[1]]
    testCoexpMat = purrr::keep(coexpMat, names(coexpMat) != cohortName)[[1]]
    refModuleList = modList
    
    innerConsvStat = plyr::ldply(refModuleList,
                                 .fun = function(x, refCoexpMat, testCoexpMat, modPreservStats){
                                   tmp = modPreservStats(refCoexpMat, testCoexpMat, x) %>%
                                     tidyr::gather(Feature, Value)
                                   
                                   tmp.rnd = purrr::map_df(1:100, .f = function(nperm, refModule, refCoexpMat, testCoexpMat){
                                     rndModule = sample(rownames(refCoexpMat), length(refModule))
                                     modPreservStats(refCoexpMat, testCoexpMat, rndModule) 
                                   },
                                   x, refCoexpMat, testCoexpMat, .id = 'nperm') %>% 
                                     tidyr::gather(Feature, Value, -nperm) %>% 
                                     dplyr::group_by(Feature) %>% 
                                     dplyr::summarise(mn = mean(Value, na.rm = T), 
                                                      sd = sd(Value, na.rm = T)) %>% 
                                     dplyr::left_join(tmp) %>% 
                                     dplyr::mutate(Z = (Value-mn)/sd)
                                   
                                   return(cbind(tmp.rnd %>%
                                                  dplyr::select(Feature, Value) %>%
                                                  tidyr::spread(Feature, Value),
                                                tmp.rnd %>% 
                                                  dplyr::mutate(Feature = paste('Z', Feature, sep = '.')) %>%
                                                  dplyr::select(Feature, Z) %>%
                                                  tidyr::spread(Feature, Z)))
                                  },
                                 refCoexpMat, testCoexpMat, modPreservStats,
                                 .id = 'ModuleName') %>%
      dplyr::mutate(reference = cohortName)
    
    return(innerConsvStat)
  }, 
  coexpMat) %>%
      dplyr::mutate(subType = subType)
  
  fwrite(consvStat, file = paste0(subType,'.tsv'), sep = '\t')
  return(consvStat)
}, cor.mat.ids)
```
### Rank conserved modules (per subtype across cohort)
```{r rnk.mdl}
conserved.mod.cohort = plyr::dlply(conservation.stats.cohort, .(subType), .fun = function(cnsvStat){
  cnsvStat = cnsvStat %>%
    dplyr::group_by(subType, reference, ModuleName) %>%
    dplyr::mutate(Z.dens = median(Z.mean.Cor, Z.mean.kIM, na.rm = T),
                  Z.con = median(Z.cor.adj, Z.cor.kIM, na.rm = T),
                  Z.summary = (Z.dens + Z.con)/2) %>%
    dplyr::arrange(desc(abs(Z.summary))) %>%
    dplyr::mutate(Cohort.ModuleName = paste(reference, ModuleName, sep = '.'))
  
  p = ggplot(cnsvStat, aes(x = Cohort.ModuleName, y = Z.summary, color = reference)) 
  p = p + geom_point() + theme(axis.text.x = element_blank(), legend.position = 'top')  
  p = p + geom_hline(yintercept = 2, color = 'blue',linetype="dashed")
  p = p + geom_hline(yintercept = 10, color = 'purple',linetype="dashed")
  p = p + ggtitle(unique(cnsvStat$subType)) + xlab('Module Name') + ylab('Z summary')
  
  cnsvStat = cnsvStat %>%
    dplyr::filter(Z.summary >= 10) %>%
    plyr::dlply(.(reference), .fun = function(x){unique(x$ModuleName)})
  
  return(c(list(p = p), cnsvStat))
})

ggpubr::ggarrange(plotlist = lapply(conserved.mod.cohort, function(x){x$p}), nrow = 2, ncol = 2)
```
### Get preserved/unpreserved modules between sub-types in each cohort
```{r presv.sub.type}
conservation.stats.subtypes = mod %>%
  purrr::imap_dfr(.f = function(innerMod, cohortName, corMatIds){
    subTypes = combn(names(innerMod), 2) %>% t 
    subTypes1 = subTypes[,c(2,1)]
    subTypes = rbind(subTypes,subTypes1) %>% unique() %>% as.data.frame()
    colnames(subTypes) = c('reference', 'against')
    subTypes = subTypes %>% dplyr::filter(reference == 'control' || against == 'control')
    
    coexpMat = corMatIds[[cohortName]] %>%
      plyr::llply(.fun = function(id){
        synGet(as.character(id))@filePath %>%
          fread(data.table = FALSE, header = T) %>%
          igraph::graph_from_data_frame(directed = FALSE) %>%
          igraph::as_adjacency_matrix(attr = 'rho', type = 'both') %>%
          as.matrix() %>%
          return()
      })
    
    cnsvStats = plyr::ddply(subTypes, .(reference, against), .fun = function(x, coexpMat, innerMod){
      refCoexpMat = coexpMat[[x$reference]]
      testCoexpMat = coexpMat[[x$against]]
      refModuleList = innerMod[[x$reference]]
      
      innerConsvStat = plyr::ldply(refModuleList,
                                   .fun = function(x, refCoexpMat, testCoexpMat, modPreservStats){
                                     tmp = modPreservStats(refCoexpMat, testCoexpMat, x) %>%
                                       tidyr::gather(Feature, Value)
                                     
                                     tmp.rnd = purrr::map_df(1:100, .f = function(nperm, refModule,
                                                                                  refCoexpMat, testCoexpMat){
                                       rndModule = sample(rownames(refCoexpMat), length(refModule))
                                       modPreservStats(refCoexpMat, testCoexpMat, rndModule) 
                                      },
                                      x, refCoexpMat, testCoexpMat, .id = 'nperm') %>% 
                                       tidyr::gather(Feature, Value, -nperm) %>% 
                                       dplyr::group_by(Feature) %>% 
                                       dplyr::summarise(mn = mean(Value, na.rm = T), 
                                                        sd = sd(Value, na.rm = T)) %>% 
                                       dplyr::left_join(tmp) %>% 
                                       dplyr::mutate(Z = (Value-mn)/sd)
                                     
                                     return(cbind(tmp.rnd %>%
                                                    dplyr::select(Feature, Value) %>%
                                                    tidyr::spread(Feature, Value),
                                                  tmp.rnd %>% 
                                                    dplyr::mutate(Feature = paste('Z', Feature, sep = '.'))
                                                  %>%
                                                    dplyr::select(Feature, Z) %>%
                                                    tidyr::spread(Feature, Z)))
                                     },
                                   refCoexpMat, testCoexpMat, modPreservStats,
                                   .id = 'ModuleName') %>%
        dplyr::mutate(reference = x$reference, against = x$against)
      
      return(innerConsvStat)
    }, coexpMat, innerMod) %>%
      dplyr::mutate(cohort = cohortName)
    
  fwrite(cnsvStats, file = paste0(cohortName, '.tsv'), sep = '\t')
  return(cnsvStats)
}, cor.mat.ids)
```
### Rank conserved modules (per cohort against subtypes)
```{r rnk.mdl1}
conserved.mod.subtypes = plyr::dlply(conservation.stats.subtypes, .(subType), .fun = function(cnsvStat){
  cnsvStat = cnsvStat %>%
    dplyr::group_by(subType, reference, ModuleName) %>%
    dplyr::mutate(Z.dens = median(Z.mean.Cor, Z.mean.kIM, na.rm = T),
                  Z.con = median(Z.cor.adj, Z.cor.kIM, na.rm = T),
                  Z.summary = (Z.dens + Z.con)/2) %>%
    dplyr::arrange(desc(abs(Z.summary))) %>%
    dplyr::mutate(Cohort.ModuleName = paste(reference, ModuleName, sep = '.'))
  
  p = ggplot(cnsvStat, aes(x = Cohort.ModuleName, y = Z.summary, color = reference)) 
  p = p + geom_point() + theme(axis.text.x = element_blank(), legend.position = 'top')  
  p = p + geom_hline(yintercept = 2, color = 'blue',linetype="dashed")
  p = p + geom_hline(yintercept = 10, color = 'purple',linetype="dashed")
  p = p + ggtitle(unique(cnsvStat$subType)) + xlab('Module Name') + ylab('Z summary')
  
  cnsvStat = cnsvStat %>%
    dplyr::filter(Z.summary >= 10) %>%
    plyr::dlply(.(reference), .fun = function(x){unique(x$ModuleName)})
  
  return(c(list(p = p), cnsvStat))
})

ggpubr::ggarrange(plotlist = lapply(conserved.mod.cohort, function(x){x$p}), nrow = 2, ncol = 2)
```
Relationship between cohorts 1 and 2 (enrichment analysis signifying the overlap between the modules)
```{r cons.mod.ovlp, fig.height=10, fig.width=20}
conserved.mod.clust = purrr::imap(conserved.mod.cohort, .f = function(x, nx, ovlpStats){
  tmp = dplyr::filter(ovlpStats[[nx]], 
                      ModuleName_c1 %in% x$cohort1,
                      ModuleName_c2 %in% x$cohort2) %>%
    dplyr::select(ModuleName_c1, ModuleName_c2, Odds.Ratio, FDR) %>%
    dplyr::mutate(ModuleName_c1 = paste('cohort1', ModuleName_c1, sep = '.'),
                  ModuleName_c2 = paste('cohort2', ModuleName_c2, sep = '.')) %>%
    unique()
  OR = igraph::graph_from_data_frame(tmp, directed = FALSE) %>%
    igraph::as_adj(type = 'both', attr = 'Odds.Ratio') %>%
    as.matrix()
  OR = OR[tmp$ModuleName_c1,tmp$ModuleName_c2]
  FDR = igraph::graph_from_data_frame(tmp, directed = FALSE) %>%
    igraph::as_adj(type = 'both', attr = 'FDR') %>%
    as.matrix()
  FDR = FDR[tmp$ModuleName_c1,tmp$ModuleName_c2]
  OR[FDR>=0.05] = 0
  
  h = ComplexHeatmap::Heatmap(log10(OR+0.1),
                              col = circlize::colorRamp2(c(0,log10(max(OR))),c('white','red')),
                              name = 'log10(Odds Ratio)')
  # ,
  #                             cell_fun = function(j, i, x, y, width, height, fill) {
  #                               grid.text(format(FDR[i, j], digits = 2),x, y, gp = gpar(fontsize = 10))
  #                             })
  return(h)
}, ovlp.stats)
conserved.mod.clust$control
conserved.mod.clust$coronal
conserved.mod.clust$metopic
conserved.mod.clust$sagittal
```
```{r consv.sts.subtypes}
tmp = conservation.stats.subtypes %>%
  dplyr::group_by(cohort, reference, against) %>%
  dplyr::mutate(Z.dens = median(Z.mean.Cor, Z.mean.kIM, na.rm = T),
                Z.con = median(Z.cor.adj, Z.cor.kIM, na.rm = T),
                Z.summary = (Z.dens + Z.con)/2) %>%
  dplyr::arrange(desc(abs(Z.summary)))

p = purrr::map(c('coronal', 'sagittal', 'metopic'), .f = function(x, tmp){
  p = tmp %>% 
    dplyr::filter(reference %in% c('control', x),
                  against %in% c('control', x)) %>%
    ggplot(aes(x = ModuleName, y = Z.summary, color = reference)) + geom_point() + facet_grid(.~cohort)
  p = p + theme(axis.text.x = element_blank(), legend.position = 'top') 
  p = p + geom_hline(yintercept = 2, color = 'blue', linetype = 'dashed')
  p = p + geom_hline(yintercept = 10, color = 'purple', linetype = 'dashed')
  return(p)
}, tmp)
ggpubr::ggarrange(plotlist = p, ncol = 1, nrow = 3)

p = purrr::map(c('sagittal', 'metopic'), .f = function(x, tmp){
  p = tmp %>% 
    dplyr::filter(reference %in% c('coronal', x),
                  against %in% c('coronal', x)) %>%
    ggplot(aes(x = ModuleName, y = Z.summary, color = reference)) + geom_point() + facet_grid(.~cohort)
  p = p + theme(axis.text.x = element_blank(), legend.position = 'top') 
  p = p + geom_hline(yintercept = 2, color = 'blue', linetype = 'dashed')
  p = p + geom_hline(yintercept = 10, color = 'purple', linetype = 'dashed')
  return(p)
}, tmp)
ggpubr::ggarrange(plotlist = p, ncol = 1, nrow = 2)

p = purrr::map(c('metopic'), .f = function(x, tmp){
  p = tmp %>% 
    dplyr::filter(reference %in% c('sagittal', x),
                  against %in% c('sagittal', x)) %>%
    ggplot(aes(x = ModuleName, y = Z.summary, color = reference)) + geom_point() + facet_grid(.~cohort)
  p = p + theme(axis.text.x = element_blank(), legend.position = 'top') 
  p = p + geom_hline(yintercept = 2, color = 'blue', linetype = 'dashed')
  p = p + geom_hline(yintercept = 10, color = 'purple', linetype = 'dashed')
  return(p)
}, tmp)
ggpubr::ggarrange(plotlist = p, ncol = 1, nrow = 1)
```

### Identify pathways that are conserved in each subtype
```{r path.cons}
path.enrich = fread(synGet('syn11924793')@filePath)

cons.path = purrr::imap(conserved.mod, function(x, nx){
  x$mod.c1 %>% 
    as.character() %>% 
    unique() %>% 
    data.frame(ModuleName = .) %>%
    dplyr::mutate(cohort = 'cohort1',
                  subtype = nx) %>%
    rbind(x$mod.c2 %>% 
    as.character() %>% 
    unique() %>% 
    data.frame(ModuleName = .) %>%
    dplyr::mutate(cohort = 'cohort2',
                  subtype = nx))
}) %>%
  rbindlist() %>%
  dplyr::left_join(path.enrich %>%
                     tidyr::separate(.id, c('cohort', 'subtype', 'ModuleName'), sep = '\\.')) %>%
  plyr::ddply(.(cohort, subtype, Category), .fun = function(x){
    x$fdr = p.adjust(x$pval, method = 'fdr')
    return(x)
  }) 


```

