---
title: "MEGENA module conservation analysis between cohort 1 and 2"
author: "Thanneer Perumal"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
### Load libraries
```{r libs, echo=FALSE, warning=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get it from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(plyr)
library(tidyverse)

library(synapser)
library(knitr)
library(githubr)
library(ComplexHeatmap)
library(grid)

synLogin()

library(foreach)
library(doParallel)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```
### Get synapse and github specific parameters
```{r synapse.parameters, include=FALSE}
# Synapse parameters
PARENT_ID = 'syn11638075'

# Github commit for the source code
thisFileName <- 'moduleConservationAnalysisC1andC2.Rmd'
thisRepo <- getRepo(repository = "th1vairam/CRANIO", ref="branch", refName='moduleAnalysis')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=thisFileName)
```
### Get modules and conservation results from synapse
```{r syn.get}
# oad module membership from synapse
mod = c(cohort1 = 'syn11635629', cohort2 = 'syn11635905') %>%
  purrr::map(.f =function(foldId){
   synapser::synQuery(paste0('select * from folder where parentId =="', foldId, '"'))$results %>%
      purrr::map_dfr(.f = function(x){as.data.table(x)}) %>%
      plyr::dlply(.(folder.name), .fun = function(x){
        tmp1 = synapser::synQuery(paste0('select * from file where parentId =="', x$folder.id, '"'))$results %>%
          purrr::map_dfr(.f = function(x){as.data.table(x)}) %>%
          dplyr::filter(file.name == 'output.RData')
        load(synGet(as.character(tmp1$file.id))$path)
        
        # Remove small or big modules
        mod = output$modules
        sz = sapply(mod, length)
        mod = mod[sz >= 30 & sz <= 2000]
        return(mod)
      })
  })

# Load module conservation results from synapse
load(synapser::synGet('syn11967846')$path)

# Load enrichment results of modules from synapse
enrich.results = synapser::synGet('syn11924793')$path %>%
  fread(data.table = F) %>%
  tidyr::separate(.id, c('Cohort', 'SubType', 'ModuleName'), sep = '\\.') %>%
  dplyr::group_by(Cohort, SubType, ModuleName, Category) %>%
  dplyr::mutate(FDR = p.adjust(pval, method = 'BH'))
```

### Filter modules that are conserved between cohort 1 and 2
```{r mdl.cnsrv, fig.height=4, fig.width=10}
conservation.stats.cohort = conservation.stats.cohort %>%
  dplyr::group_by(reference, subType, ModuleName) %>%
  dplyr::mutate(Z.dens = median(Z.mean.Cor, Z.mean.kIM, na.rm = T),
                  Z.con = median(Z.cor.adj, Z.cor.kIM, na.rm = T),
                  Z.summary = (Z.dens + Z.con)/2) %>%
  dplyr::arrange(desc(abs(Z.summary)))

p = ggplot(conservation.stats.cohort, aes(x = ModuleName, y = Z.summary)) + geom_point()
p = p + theme_bw() %+replace% theme(axis.text.x = element_blank())
p = p + geom_hline(yintercept = 3, color = 'red', linetype = 'dashed')
p = p + geom_hline(yintercept = 8, color = 'darkgreen', linetype = 'dashed')
p = p + facet_grid(reference ~ . + subType) +  ggtitle('Module conservation between cohorts 1 and 2')
p

conserved.mod = dplyr::filter(conservation.stats.cohort, Z.summary >= 8)
kable(conserved.mod)

conserved.mod %>%
  dplyr::group_by(subType, reference) %>%
  dplyr::summarise(ModuleName = paste(unique(ModuleName), collapse = ', ')) %>%
  tidyr::spread(reference, ModuleName)
```
### Find overlap/enrichment between conserved modules in cohort 1 and 2
```{r mdl.ovlp.c1.c2, fig.height=8, fig.width=12}
bck.genes = unlist(unlist(unlist(mod))) %>% unique()
enrichment.bw.cohort = conserved.mod %>%
  plyr::ddply(.(subType), .fun = function(x, mod, bckGenes){
    c1 = dplyr::filter(x, reference == 'cohort1') %>% 
      dplyr::select(ModuleName) %>%
      unlist() %>%
      unique()
    c1 = mod$cohort1[[unique(x$subType)]][c1]
    
    c2 = dplyr::filter(x, reference == 'cohort2') %>% 
      dplyr::select(ModuleName) %>%
      unlist() %>%
      unique()
    c2 = mod$cohort2[[unique(x$subType)]][c2]
    
    tmp = purrr::cross2(c1, c2) %>%
      purrr::map(.f = function(lm, bckGenes){
        CovariateAnalysis::fisherEnrichment(lm[[1]], lm[[2]], bckGenes) 
      }, bckGenes) %>%
      dplyr::bind_rows() %>%
      cbind(expand.grid(names(c1), names(c2)),.) %>%
      dplyr::rename(Cohort1 = Var1, Cohort2 = Var2)
  }, mod, bck.genes) %>%
  dplyr::select(-Genes) %>%
  dplyr::group_by(subType) %>%
  dplyr::mutate(fdr = p.adjust(pval, method = 'BH'))

ovlp.plots.bw.cohort = enrichment.bw.cohort %>%
  dplyr::ungroup() %>%
  plyr::dlply(.(subType), .fun = function(x){
    or = x %>%
      dplyr::select(Cohort1, Cohort2, Odds.Ratio) %>%
      tidyr::spread(Cohort2, Odds.Ratio) %>%
      as.data.frame()
    rownames(or) = or$Cohort1
    or$Cohort1 = NULL
    
    fdr = x %>%
      dplyr::select(Cohort1, Cohort2, fdr) %>%
      tidyr::spread(Cohort2, fdr) %>%
      as.data.frame()
    rownames(fdr) = fdr$Cohort1
    fdr$Cohort1 = NULL
    
    or[fdr >= 0.05] = 0
    
    hm = ComplexHeatmap::Heatmap(or, name = 'OddsRatio', column_title = unique(x$subType))
    g = grid.grabExpr(draw(hm))
})
ggpubr::ggarrange(plotlist = ovlp.plots.bw.cohort, ncol=2, nrow = 2)
```

### Filter modules from cohort 1 that are conserved in 2, and plot between subTypes
```{r mdl.cnsrvd.c1, fig.height=8, fig.width=10}
conservation.stats.subtypes = conservation.stats.subtypes %>%
  dplyr::group_by(reference, against, cohort, ModuleName) %>%
  dplyr::mutate(Z.dens = median(Z.mean.Cor, Z.mean.kIM, na.rm = T),
                  Z.con = median(Z.cor.adj, Z.cor.kIM, na.rm = T),
                  Z.summary = (Z.dens + Z.con)/2) %>%
  dplyr::arrange(desc(abs(Z.summary)))

conserved.mod2 = conservation.stats.subtypes %>%
  dplyr::select(ModuleName, reference, against, cohort, Z.summary) %>%
  dplyr::inner_join(conserved.mod %>%
                      dplyr::select(ModuleName, reference, subType) %>%
                      dplyr::rename(cohort = reference, reference = subType))

p = ggplot(conserved.mod2, aes(x = ModuleName, y = Z.summary, color = cohort)) + geom_point()
p = p + theme_bw() %+replace% theme(axis.text.x = element_blank())
p = p + geom_hline(yintercept = 3, color = 'red', linetype = 'dashed')
p = p + geom_hline(yintercept = 8, color = 'darkgreen', linetype = 'dashed')
p = p + facet_grid(reference ~ . + against) + ggtitle('Module conservation between subtypes')
p
```
Modules conserved in cohort 1 and 2, and not conserved between sub types
```{r anlz1, results='asis'}
tmp = conserved.mod2 %>%
  dplyr::filter(Z.summary <= 3) %>%
  dplyr::group_by(cohort, reference, against) %>%
  dplyr::summarise(ModuleName = paste(unique(ModuleName), collapse = ', ')) %>%
  tidyr::spread(against, ModuleName)
kable(tmp)
```
## Annotate modules
```{r de.enrichment}
de.gs = c(cohort1 = 'syn8555303', cohort2 = 'syn8555329') %>%
  purrr::map(.f = function(id){ fread(synapser::synGet(id)$path, data.table = F) }) %>%
  dplyr::bind_rows(.id = 'cohort') %>%
  dplyr::filter(Direction != 'NONE') %>%
  plyr::dlply(.(cohort, Model, Comparison, Direction), .fun = function(x){
    unique(x$ensembl_gene_id)
  })
de.gs = de.gs[sapply(de.gs, length) >= 20]

enrich.de = plyr::ldply(mod, .fun = function(c1, deGS, bckGenes){
  plyr::ldply(c1, .fun = function(s1, deGS, bckGenes){
    plyr::ldply(s1, .fun = function(modMembers, deGS, bckGenes){
      plyr::ldply(deGS, .fun = function(gs, modMembers, bckGenes){
        CovariateAnalysis::fisherEnrichment(modMembers, gs, bckGenes)
      }, modMembers, bckGenes, .id = 'SetName')
    }, deGS, bckGenes, .id = 'ModuleName')
  }, deGS, bckGenes, .id = 'subType')
},de.gs, bck.genes, .id = 'cohort') %>%
  dplyr::rename(Cohort = cohort, SubType = folder.name) %>%
  dplyr::group_by(Cohort, SubType) %>%
  dplyr::mutate(FDR = p.adjust(pval, method = 'BH'),
                Category = 'DifferentialExpression')

enrich = list(enrich.results, enrich.de) %>%
  data.table::rbindlist(use.names = T, fill = T)
```

Annotate modules with pathway enrichment
```{r annot.module}
conserved.mod.annotation = conserved.mod2 %>%
  dplyr::filter(Z.summary <= 3) %>%
  dplyr::rename(Cohort = cohort, SubType = reference) %>%
  dplyr::left_join(enrich.results %>%
                     dplyr::filter(FDR <= 0.05, Odds.Ratio >= 2)) %>%
  dplyr::group_by(Cohort, SubType, against, ModuleName) %>%
  # dplyr::top_n(10, Odds.Ratio) %>%
  dplyr::mutate(SetName = gsub('GO_Biological_Process.','',SetName)) %>%
  dplyr::summarise(SetName = paste(unique(SetName), collapse = ', ')) %>%
  kable()
```